---
title: "Artificial Neural Networtk"
subtitle: "[Public GitHub Repository](https://github.com/paACode/publictransport_winterbottlenecks_palest)"
author:  "Pascal"
date: "05.04.2025"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r, echo=FALSE , include=FALSE}
## Document related Settings and Libs
library(knitr) 
library(details)    # allows collapsible code blocks
library(gt)         # formats tables nicely
library(scales)   

## general language settings
Sys.setenv(LANG = "en")

## Default Settings for R-Chunks
opts_chunk$set(echo = FALSE ,
               include = FALSE,
               comment = NA,
               eval = TRUE,
               message = FALSE,
               warning = FALSE)
```

```{r Imports}

library(dplyr)
library(nnet)
library(gamlss.add)
library(dplyr)
library(ggplot2)
library(caret)
theme_set(theme_bw())

```

This chapter investigates the predictability of train cancellations without prior exploratory data analysis. It serves as an initial approach to gain a first understanding of the underlying data. More detailed analyses using simpler and more interpretable models will follow in subsequent chapters. Nevertheless, an artificial neural network (ANN) enables the construction of a predictive model without extensive knowledge of the data structure and can still provide initial indications as to whether the data contains a meaningful signal.

So we directly jump in an try to predict if a train is cancelled based on all available variables except the ones which already reveal if a train is cancelled or not which are:
- AN_PROGNOSE and AB_PROGNOSE: Is always NA when train is cancelled
- AN_PRONOSE_STATUS and AB_PROGNOSE_STATUS : Is always "UNBEKANNT" when cancelled
- ABFAHRTDELAY_sec", "ABFAHRTDELAY_min","ANKUNFTDELAY_sec", "ANKUNFTDELAY_min" can in consequence also not be used as 

```{r Read Data}
#### Read in Data ####----------------------------------------------------------
v.to.drop <- c("AN_PROGNOSE", "AN_PROGNOSE_STATUS", "AB_PROGNOSE",
                  "ABFAHRTDELAY_sec", "ABFAHRTDELAY_min",
                  "ANKUNFTDELAY_sec", "ANKUNFTDELAY_min")

d.ann<- read.csv("zentrahlbahn_final.csv") %>% 
  select(-all_of(v.to.drop))


```

```{r Ratio Cancelled}
d.ann.cancelled <- d.ann %>% 
  filter(FAELLT_AUS_TF == TRUE)
cancelled <- nrow(d.ann.cancelled)
tot <- nrow(d.ann)
r.cancelled <- cancelled/tot
rm(d.ann.cancelled)
```


The first step is to see how many trains are cancelled.  It is very intersting that `r percent(r.cancelled)` of all observations are cancelled

For the data preparation we need to make sure all the variables are either numerical or factors which can be one-hot encoded. Therefore we drop the times
"ANKUNFTSZEIT" and "ABFAHRTSZEIT" becuase they would result in a huge amount of levels.


```{r Perpare Data}
convert_all_possible_to_factor <- function(df) {
  df[] <- lapply(df, function(col) {
    if (is.character(col) || is.logical(col)) {
      as.factor(col)
    } else {
      col
    }
  })
  return(df)
}
drop_factors_with_less_than_two_levels <- function(df) {
  # Identify factors with fewer than two levels (0 or 1)
  small_level <- sapply(df, function(x) is.factor(x) && length(levels(x)) < 2)
  
  # Remove those columns
  df_clean <- df[, !small_level, drop = FALSE]
  
  return(df_clean)
}

d.ann <- convert_all_possible_to_factor(d.ann)
d.ann <- drop_factors_with_less_than_two_levels(d.ann)
d.ann <- d.ann %>% select(!ANKUNFTSZEIT) #Remove Time 
d.ann <- d.ann %>% select(!ABFAHRTSZEIT) #Remove Time 

#Create Dummy Variables 
dummies <- dummyVars(~ ., data = d.ann)
data_dummies <- predict(dummies, newdata = d.ann)

```


`



































```{r Inspect Variables}
#### Read in Data ####----------------------------------------------------------

stops <-  unique(d.ann.cancelled$HALTESTELLEN_NAME)
betriebstag <-  unique(d.ann.cancelled$BETRIEBSTAG)
table(d.ann.cancelled$BETRIEBSTAG) # There seem to be some days which are more prone to delay, from 25 on food again --> 

table(d.ann.cancelled$VERKEHRSMITTEL_TEXT) # A lot of R trains are cancelled
table(d.ann.cancelled$ZUSATZFAHRT_TF) #Zusatzfahrt are not enough

table(d.ann.cancelled$LINIEN_TEXT) # R70 and R71 would be very good predictors

table(d.ann.cancelled$HALTESTELLEN_NAME) # It would be interesting how good it performs after we remove data from Brienz and stuff

table(d.ann.cancelled$START)
table(d.ann.cancelled$ZIEL)


d.ann.reduced <- d.ann.cancelled %>% filter(!LINIEN_TEXT %in% c("R70", "R71"))

nrow(d.ann.reduced)

table(d.ann.reduced$HALTESTELLEN_NAME)



```

