---
title: "dataset_modification"
output: html_document
---

# Load CSV from the current working directory

raw_data \<- read.csv("your_file.csv", header = TRUE, stringsAsFactors = FALSE)

``` {r Load CSV}
# Load CSV from the current working directory
raw_imported_data <- read.csv("zentrahlbahn_nov_dataset.csv", header = TRUE, stringsAsFactors = FALSE)
```


``` {r modify data}
library(dplyr)
library(lubridate)

data_processed <- raw_imported_data %>%
  mutate(
    BETRIEBSTAG = as.Date(BETRIEBSTAG, format = "%d.%m.%Y"),
    BETREIBER_ABK = as.factor(BETREIBER_ABK),
    BETREIBER_NAME = as.factor(BETREIBER_NAME),
    PRODUKT_ID = as.factor(PRODUKT_ID),
    LINIEN_TEXT = as.factor(LINIEN_TEXT),
    VERKEHRSMITTEL_TEXT = as.factor(VERKEHRSMITTEL_TEXT),
    HALTESTELLEN_NAME = as.factor(HALTESTELLEN_NAME),
    # Convert empty strings to NA for time columns BEFORE conversion:
    ANKUNFTSZEIT = na_if(ANKUNFTSZEIT, ""),
    AN_PROGNOSE   = na_if(AN_PROGNOSE, ""),
    ABFAHRTSZEIT  = na_if(ABFAHRTSZEIT, ""),
    AB_PROGNOSE   = na_if(AB_PROGNOSE, ""),
    # Now convert to POSIXct
    ANKUNFTSZEIT = as.POSIXct(ANKUNFTSZEIT, format = "%d.%m.%Y %H:%M"),
    AN_PROGNOSE   = as.POSIXct(AN_PROGNOSE, format = "%d.%m.%Y %H:%M:%S"),
    ABFAHRTSZEIT = as.POSIXct(ABFAHRTSZEIT, format = "%d.%m.%Y %H:%M"),
    AB_PROGNOSE   = as.POSIXct(AB_PROGNOSE, format = "%d.%m.%Y %H:%M:%S"),
    # Calculate delays: SOLL minus IST
    delay_sec = as.numeric(difftime(AN_PROGNOSE, ANKUNFTSZEIT, units = "secs")),
    delay_min = round(delay_sec / 60, 1)
  ) %>%
  # Group by train ID (FAHRT_BEZEICHNER)
  group_by(FAHRT_BEZEICHNER) %>%
  mutate(
    # For START, take the HALTESTELLEN_NAME where both ANKUNFTSZEIT and AN_PROGNOSE are missing
    START = first(as.character(HALTESTELLEN_NAME)[is.na(ANKUNFTSZEIT) & is.na(AN_PROGNOSE)]),
    # For ZIEL, take the HALTESTELLEN_NAME where both ABFAHRTSZEIT and AB_PROGNOSE are missing
    ZIEL  = first(as.character(HALTESTELLEN_NAME)[is.na(ABFAHRTSZEIT) & is.na(AB_PROGNOSE)])
  ) %>%
  ungroup()



avg_delay_summary_full <- data_processed %>%
  group_by(FAHRT_BEZEICHNER) %>%
  summarise(
    # For all other columns, take the first value in the group
    across(-c(delay_sec, delay_min), first),
    avg_delay_sec = mean(delay_sec, na.rm = TRUE),
    avg_delay_min = round(mean(delay_sec, na.rm = TRUE) / 60, 1)
  ) %>%
  ungroup()



```