---
title: "Artificial Neural Networtk"
subtitle: "[Public GitHub Repository](https://github.com/paACode/publictransport_winterbottlenecks_palest)"
author:  "Pascal"
date: "05.04.2025"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r, echo=FALSE , include=FALSE}
## Document related Settings and Libs
library(knitr) 
library(details)    # allows collapsible code blocks
library(gt)         # formats tables nicely
library(scales)   

## general language settings
Sys.setenv(LANG = "en")

## Default Settings for R-Chunks
opts_chunk$set(echo = FALSE ,
               include = FALSE,
               comment = NA,
               eval = TRUE,
               message = FALSE,
               warning = FALSE)
```

```{r Imports}

library(dplyr)
library(nnet)
library(gamlss.add)
library(dplyr)
library(ggplot2)
library(caret)
theme_set(theme_bw())

```

# A First Look: Predicting Train Cancellations with an Artificial Neural Network

This chapter investigates the predictability of train cancellations without performing prior exploratory data analysis (EDA). It serves as an initial attempt to understand the underlying data. More detailed analyses using simpler and more interpretable models will follow in subsequent chapters.

An artificial neural network (ANN) enables the construction of a predictive model without requiring extensive knowledge of the data structure. It can provide early indications of whether the data contains a meaningful signal.

Therefore, the approach directly attempts to predict whether a train is cancelled, using all available variables, except those that explicitly reveal cancellation status. These are:


- `AN_PROGNOSE` and `AB_PROGNOSE`: Always `NA` when a train is cancelled
- `AN_PROGNOSE_STATUS` and `AB_PROGNOSE_STATUS`: Always `"UNBEKANNT"` when cancelled
- All derived columns such as:
  - `ABFAHRTDELAY_sec`, `ABFAHRTDELAY_min`
  - `ANKUNFTDELAY_sec`, `ANKUNFTDELAY_min`
  - `RUSH_HOUR`, `TAGESZEIT`, `DELAY_CATEGORY`
  - `ZIEL`, `START`
  
These variables are excluded to avoid data leakage.


```{r Read Data}
#### Read in Data ####----------------------------------------------------------
v.to.drop <- c("AN_PROGNOSE", "AN_PROGNOSE_STATUS", "AB_PROGNOSE",
               "AB_PROGNOSE_STATUS",
               "ABFAHRTDELAY_sec", "ABFAHRTDELAY_min",
               "ANKUNFTDELAY_sec", "ANKUNFTDELAY_min", "TAGESZEIT", 
               "RUSH_HOUR", "Delay_Category", "ZIEL", "START", "FAHRT_BEZEICHNER"
)

d.ann<- read.csv("../data/zentrahlbahn_final.csv") %>% 
  select(-all_of(v.to.drop), -starts_with("w_")) # Also drop all weather data


```

```{r Ratio Cancelled}
d.ann.cancelled <- d.ann %>% 
  filter(FAELLT_AUS_TF == TRUE)
cancelled <- nrow(d.ann.cancelled)
tot <- nrow(d.ann)
r.cancelled <- cancelled/tot
rm(d.ann.cancelled)
```


## Data Preparation

As a first step, the distribution of the response variable is examined. An initial analysis indicates that `r percent(r.cancelled)` of the trains in the dataset are cancelled. This was taken into account during data preparation. For the ANN to work, the following steps were performed:

1. Convert character and logical variables to factors  
2. Remove factors with fewer than two levels (contain no information)  
3. Use only factor variables to keep the ANN simple  
4. Drop time columns because they include a large number of levels  
5. Perform one-hot encoding  
6. Stratify train and test data  
7. Downsample train data to ensure an equal number of cancelled and uncancelled trains  



```{r Perpare Data}
convert_all_possible_to_factor <- function(df) {
  df[] <- lapply(df, function(col) {
    if (is.character(col) || is.logical(col)) {
      as.factor(col)
    } else {
      col
    }
  })
  return(df)
}
drop_factors_with_less_than_two_levels <- function(df) {
  # Identify factors with fewer than two levels (0 or 1)
  small_level <- sapply(df, function(x) is.factor(x) && length(levels(x)) < 2)
  
  # Remove those columns
  df_clean <- df[, !small_level, drop = FALSE]
  
  return(df_clean)
}
drop_non_factors <- function(df) {
  df[, sapply(df, is.factor), drop = FALSE]
}

d.ann <- convert_all_possible_to_factor(d.ann)
d.ann <- drop_factors_with_less_than_two_levels(d.ann)
d.ann <- drop_non_factors(d.ann) 
d.ann <- d.ann %>% select(!ANKUNFTSZEIT) #Remove Time 
d.ann <- d.ann %>% select(!ABFAHRTSZEIT) #Remove Time 

d.ann.rv <-  d.ann %>% select(FAELLT_AUS_TF) # Response Variable
d.ann.pred <-  d.ann %>% select(BETRIEBSTAG, LINIEN_TEXT,HALTESTELLEN_NAME) 
# BEtriebstag, LinienText and Haltestellenname are the predictors for the model


#Create Dummy Variables 
dummies <- dummyVars(~ ., data = d.ann.pred)
data_dummies <- predict(dummies, newdata = d.ann.pred)

# Add one hot encoded predictors to RV
d.ann.rdy <- bind_cols(d.ann.rv, data_dummies)

```


```{r Test and Train Data, echo =TRUE}
set.seed(123)
train_index <- createDataPartition(d.ann.rdy$FAELLT_AUS_TF, p = 0.8, list = FALSE)

train <- d.ann.rdy[train_index, ]
test <- d.ann.rdy[-train_index, ]

# Make sure there are as much Cancelled as Non Cancelled observations in Train Data
majority_class <- train[train$FAELLT_AUS_TF == FALSE, ]
minority_class <- train[train$FAELLT_AUS_TF == TRUE, ]

# Sample from majority class to match minority class size
set.seed(123)
majority_downsampled <- majority_class[sample(nrow(majority_class), nrow(minority_class)), ]

# Combine minority class with downsampled majority class
train_downsampled <- rbind(minority_class, majority_downsampled)

# Shuffle rows 
train_downsampled <- train_downsampled[sample(nrow(train_downsampled)), ]



```

## Building the Network

Building the Network included a lot of trial and error. Because for many configurations the model did not converge. In the end 3 predictors could be found which lead to quite good prediction:

- BETRIEBSTAG
- LINIEN_TEXT
- HALTESTELLEN_NAME



```{r Build the network, echo=TRUE, cache =TRUE}


set.seed(412)
fealltaus_net <- nnet(FAELLT_AUS_TF ~  ., data = train_downsampled, size=15, maxit=500, range=0.1, decay=1e-4, MaxNWts = 10000)
```

```{r Plot NET, include =TRUE}
plot(fealltaus_net)
```

## Results

The model successfully predicts train cancellations using only three categorical predictors: **BETRIEBSTAG**, **LINIEN_TEXT**, and **HALTESTELLEN_NAME**. Despite the test data containing 80% uncancelled and 20% cancelled trains, the model performs well thanks to stratification and downsampling during training. These results suggest that these predictors are promising candidates for more detailed modeling and analysis.


```{r Test the Network, include=TRUE}
test.n.result <- test %>% select(-FAELLT_AUS_TF)
pred <- predict(fealltaus_net, test.n.result, type="class")
cm_nn <- table(pred=pred, true=test$FAELLT_AUS_TF)
cm_nn

confusionMatrix(as.factor(pred), as.factor(test$FAELLT_AUS_TF))

```
